\pkg{cleanR}: Maid for Cleaning Datasets in \proglang{R}

### 2016-12-30 latex/article_vol2.tex
Documentation

\smallskip Control summarize, visualize, and check steps \\
\quad \R{mode} & What steps should be performed for each variable (out of the three possibilities \textit{summarize}, \textit{visualize}, \textit{check})? & \R{c("summarize", "visualize", "check")} (corresponding to all three) \\
\quad \R{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \R{TRUE} \\
\quad \R{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \R{Inf} \\
\quad \R{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \R{2} \\
\quad \R{twoCol} & Should the summary table and visualizations be placed side-by-side (in two columns)? & \R{TRUE} \\

@@

\subsection{Rolling out the arguments}
We begin with an example that is intended as an illustration of how \R{clean()} might
be used in the very first stages of data cleaning, where we are still uncertain
about exactly what is needed and how much times should be allocated to data
cleaning. At this stage, what is really needed, if a very rough idea of the extends of
errors in the dataset. In this scenario, we might wish to obtain a summary document in html format 
that only contains the variables with potential problems, and with a limit of
maximum 10 printed potential errors for each variable we can write
(output not shown). Also, we can add the argument \R{replace=TRUE} in
order to force \R{clean()} to overwrite any existing files produced by
\R{clean()}, as we might run this function a few times in a row before deciding
exactly which cleaning overview we want. Using the \R{toyData} dataset as a guinea pig, we type:

@@

The final rendering of the generated markdown file is controlled by
the \R{render} and \R{openResult} arguments, which both default to
\R{TRUE} \hl{Bliver det ved med at være sandt?}. \R{render} determines
if the \proglang{R} markdown file produced should be rendered using
the \pkg{rmarkdown} package and \R{openResult} decides whether the
output html or pdf file should be opened. The following command
produces an \proglang{R} markdown file containing the data cleaning
summary information, but without rendering nor opening the markdown file:

@@

We will now move on to discuss how not only the structure of the cleaning summary is manipulated, but also its very contents. This is done through the summarize-visualize-check (SVC) step, as illustrated in Figure \ref{figure:cleanStructure}. 
\pkg{cleanR} uses three different types of functions for performing these steps,
namely \code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.
By default, \R{clean()} runs selected summary, visualization and check functions on each variable in the dataset, and the exact choice of these functions depends on the classes of the variables. For instance, though detection of outlier values might be interesting for numerical variables, it holds little meaning for factor variables, and therefore, numerical and factor variables need different checks. Table~\ref{table.SVCfunctions}
lists all available SVC functions, but we can also use the
\R{allSummaryFunctions()}, \R{allVisualFunctions()}, and
\R{allCheckFunctions()} functions in \pkg{cleanR} to print overview
lists in \proglang{R}. For example, the implemented \R{summaryFunction}s are:

@@

As with \R{summaryFunction}s, a complete list of available \R{checkFunction}s is obtained by calling 
\R{allCheckFunctions()}. Note however, that \R{checkFunction}s have a usage beyond the 
\R{XXXChecks} arguments, namely in the \textit{pre-check} stage. In this stage, it is determined 
whether or not each variable is suitable for the SVC-steps. The functions used in the pre-check stage
should be \R{checkFunction}s that are applicable to all variable classes. The default pre-checks, the functions
\R{isKey} and \R{isEmpty}, check whether a variable has unique values for all observations or only a single value for 
all observations, respectively. If one of these statements are true, the variable will not be subjected to the SVC steps. 
We can allow empty variables to move on to the SVC step by only checking for keys in the pre-check step:

@@

As we saw with the \R{check()} function, the summary can be modified
by providing the relevant \R{XXXSummaries} arguments. Setting the
\R{numericSummaries} argument, we can control the summary output by
providing a vector of function names to run for a particular summary. To
only get the median, the minimum and the maximum we set
\R{numericSummaries=c("centralValue", "minMax")}:

@@

This section consists of two parts. First, we describe how custom \R{summaryFunction}s, 
\R{visualFunction}s and \R{checkFunctions} can be made, including an introduction to a 
few convenient tools available in \pkg{cleanR} for aiding SVC-function construction. 
Table \ref{table.functionTypes} serves as an overview of the internals of these three
central function types. Secondly, we turn to a 
worked example of how to use custom made functions in practice. Here,
four new SVC functions are defined and used, both interactively and in \R{clean()}.

\begin{table}[htbp]
\footnotesize
\bgroup
\def\arraystretch{1.8}% 
\begin{tabular}{L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}}
& \R{summaryFunction} & \R{visualFunction} & \R{checkFunction} \\
\hline
\vspace{0pt} Input (required) & \R{v} - a variable vector \newline \R{...}  &  \R{v} - a variable vector \newline \R{vnam} - the variable name (as character string) \newline \R{doEval} - a logical (\R{TRUE}/\R{FALSE}) controlling the output type of the function & \R{v} - a variable vector \newline \R{nMax} - an integer (or \R{Inf}), controlling how many problematic values are printed, if relevant \newline \R{...}  \\
Input (optional) &  \R{maxDecimals} - number of decimals printed in outputted numerical values.  & - &  \R{maxDecimals}  - number of decimals printed in outputted numerical values.  \\
Purpose & Describe some aspect of the variable, e.g. a central value, its dispersion or level of missingness. & Produce a distribution plot. & Check a variable for a specific issue and, if relevant, identify the values in the variable that cause the issue. \\
Output (required) & A list with entries: \newline \R{\$feature} - a label for the summary value (as character string); \newline \R{\$result} - the result of the summary (as character string) & A character string with \R{R} code for producing a plot. This code should be standalone, i.e. should include the data if necessary. & A list with entries: \newline \R{\$problem} - a logical identifying whether an issue was found; \newline \R{\$message} - a character string (possibly empty) decribing the issue that was found, properly escaped and ready for use in \R{rmarkdown} \\
Output (recommended) & A \R{summaryResult} object, i.e. an attributed list with entries \R{\$feature}, \R{\$result} and \R{\$value}, the latter being the values from \R{\$result} in their original format). & \bgroup
\def\arraystretch{1.8}% 
\hspace*{-.2cm}\begin{tabular}[t]{p{0.45\linewidth} | p{0.45\linewidth}} \vspace*{-.9cm}If \R{doEval} is \R{TRUE}: \newline A plot that is opened by the graphic device in \R{R}. & \vspace*{-.9cm}If \R{doEval} is \R{FALSE}: \newline   A text string with \R{R} code, as described above. \end{tabular} \egroup  & A \R{checkResult} object, i.e. an attributed list with entries \R{\$problem}, \R{\$message} and \R{\$problemValues}, the latter being either \R{NULL} or the problem causing values, as they were found in \R{v}, whichever is relevant. \\
Tools available for producing the function & \R{summaryResult()} & --- & \R{messageGenerator()} \newline \R{checkResult()} 
\end{tabular}
\egroup
\caption{Reference information for creating new functions to be used
  as part of the \mbox{summarize-,} visualize-, and check steps. The three
  columns correspond to each of the three function types. %\hl{I think the formatting here is still sort of chaotic. Maybe include horizontal lines everywhere? Is it allowed in JSS? It's impossible to read what's in which row right now.}
}
\label{table.functionTypes}
\end{table}

@@

Adhering to the template above is sufficient for using the freshly
made \R{mySummaryFunction()} in \R{clean()}, but we recommend
furthermore adding the new function to the overview of all summary
functions by converting it to a proper \code{summaryFunction}
object. This is done by calling the \R{summaryFunction()} creator with
the user-defined function as the first argument, and additional arguments
\R{description} (an explanatory text which will be added to the attributes of
the function), and \R{classes} (a vector of variable classes the
user-defined function is intended to be applied to, also stored as an
attribute). In other words, a call on the following form should be made:
\begin{Schunk}
\begin{Sinput}
mySummaryFunction <- summaryFunction(mySummaryFunction,
  description = "[Text describing what the summaryFunction does]",
  classes = c([vector of data types that the function is intended for]))
\end{Sinput}
\end{Schunk}
which adds the new function to the output of an
\code{allSummaryFunctions()} call. 
% One comment should be devoted to
% the two attributes of a \R{summaryFunction}. 
If the \R{description} argument is left unspecified, the name of the
function (which in this case is \R{"mySummaryFunction"}) will be filled in and
stored under the \R{description} attribute. What happens if the \R{classes} argument is
not specified depends on the type of \R{mySummaryFunction}. If
\R{mySummaryFunction} is a \R{S3} generic function with associated
methods, the call to \R{summaryFunction()} will automatically produce
a vector of the names of the classes for which the function can be
called. If \R{mySummaryFunction} is not an \R{S3} generic and
\R{classes} is left unspecified, the attribute will simply be
left empty. Note that the helper function \R{allClasses()} might be useful
for filling out the \R{classes} argument, as it simply lists all
available classes in \pkg{cleanR}:
\begin{Schunk}
\begin{Sinput}
> allClasses()
\end{Sinput}
\begin{Soutput}
[1] "character" "Date"      "factor"   "integer"   "labelled"  
[6] "logical"  "numeric"  
\end{Soutput}
\end{Schunk}
% \hl{Write something here, don't end paragraph with code. Also, maybe move the allClasses() stuff somewhere else, it doesn't really belong under this header. Not sure where to, though.}

@@

myProbValCheckFunction <- checkFunction(myProbValCheckFunction,
  description = "[Some text describing the checkFunction]",
  classes = c([the data types that this function is intended to be used for])
)
\end{Sinput}
\end{Schunk}
One comment should be devoted to the helper function, \R{messageGenerator}. 
This function's sole purpose is aiding consistent
styling of all \code{checkFunction} messages. The function simply
pastes together the \code{problemMessage} and the
\code{problemValues}, with the latter being quoted and sorted
alphabetically. If \R{nMax} is not \R{Inf}, only the first \R{nMax}
problem values will be pasted onto the message, accompanied by a
comment about how many problem values were left out (if any).  Note
that printing quotes in \pkg{rmarkdown} requires an extensive amount
of character escaping, so opting for \code{messageGenerator()} really
is the easiest solution.

@@

\hl{Jeg forstår ikke helt, hvad der var uklart her ang. messageGenerator. Jeg har nu
stort set bare byttet om på rækkefølgen af de to sidste paragraphs her, hjælper
det?}

@@

\section{Zooming in on cleaning challenges}
\label{sec:specificExamples}
\hl{Lidt poppet titel, men den gamle ("Using cleanR in specific situations")
fik mig til flabel at tænke "i modsætning til uspecifikke situationer?". Jeg 
er dog meget åben for en bedre titel.}

@@

On the other hand, if a pdf document is to be produced, a few extra
lines need to be added to the preamble of the master \pkg{rmarkdown}
document --- otherwise, the two-column layout code will produce an
error. The following is an example of how such a master document
preamble YAML might look like and how the \R{cleanR\_toyData.Rmd} file can
then be included:
\begin{Verbatim}
---
output: pdf_document
documentclass: report
header-includes:
  - \renewcommand{\chaptername}{Part}
  - \newcommand{\fullline}{\noindent\makebox[\linewidth]{\rule{\textwidth}{0.4pt}}}
  - \newcommand{\bminione}{\begin{minipage}{0.75 \textwidth}}
  - \newcommand{\bminitwo}{\begin{minipage}{0.25 \textwidth}}
  - \newcommand{\emini}{\end{minipage}}
---

```{r, child = 'cleanR_toyData.Rmd'}
```
\end{Verbatim}
%\hl{Use proper formatting here. How do we do non-R code?}

>>>>

\smallskip Control summarize, visualize, and check steps \\
\quad \R{mode} & What steps should be performed for each variable (out
                 of the three possibilities \textit{summarize},
                 \textit{visualize}, \textit{check})? &
                                                        \R{c("summarize", "visualize", "check")} (corresponding to all three steps) \\
\quad \R{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \R{TRUE} \\
\quad \R{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \R{Inf} \\
\quad \R{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \R{2} \\
\quad \R{twoCol} & Should the summary table and visualizations be placed side-by-side (in two columns)? & \R{TRUE} \\

@@

\subsection{Polishing off the arguments}
We begin with an example that is intended as an illustration of how \R{clean()} might
be used in the very first stages of data cleaning, where we are still uncertain
about exactly what is needed and how much times should be allocated to data
cleaning. At this stage, what is really needed, if a very rough idea of the extends of
errors in the dataset. In this scenario, we might wish to obtain a summary document in html format 
that only contains the variables with potential problems, and with a limit of
maximum 10 printed potential errors for each variable we can write
(output not shown). Also, we can add the argument \R{replace=TRUE} in
order to force \R{clean()} to overwrite any existing files produced by
\R{clean()}, as we might run this function a few times in a row before deciding
exactly which cleaning overview we want. Using the \R{toyData} dataset as a guinea pig, we type:

@@

The final rendering of the generated markdown file is controlled by
the \R{render} and \R{openResult} arguments, which both default to
\R{TRUE}. \R{render} determines if the \proglang{R} markdown file
produced should be rendered using the \pkg{rmarkdown} package and
\R{openResult} decides whether the output html or pdf file should be
opened. The following command produces an \proglang{R} markdown file
containing the data cleaning summary information, but without
rendering nor opening the markdown file:

@@

We will now move on to discuss how not only the structure of the
cleaning summary is manipulated, but also its very contents. This is
done through the summarize-visualize-check (SVC) step, as illustrated
in Figure \ref{figure:cleanStructure}.  \pkg{cleanR} uses three
different types of functions for performing these steps, namely
\code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.  By default, \R{clean()} runs selected summary,
visualization and check functions on each variable in the dataset, and
the exact choice of these functions depends on the classes of the
variables. For instance, though detection of outlier values might be
interesting for numerical variables, it holds little meaning for
factor variables, and therefore, numerical and factor variables need
different checks. Table~\ref{table.SVCfunctions} lists all available
summarize/visualize/check functions, but we can also use the
\R{allSummaryFunctions()}, \R{allVisualFunctions()}, and
\R{allCheckFunctions()} functions in \pkg{cleanR} to print overview
lists in \proglang{R}. For example, the implemented
\R{summaryFunction}s are:

@@

As with \R{summaryFunction}s, a complete list of available
\R{checkFunction}s is obtained by calling
\R{allCheckFunctions()}. Note however, that \R{checkFunction}s have a
usage beyond the \R{XXXChecks} arguments, namely in the
\textit{pre-check} stage. In this stage, it is determined whether or
not each variable is suitable for the summarize/visualize/check (SVC)
steps. The functions used in the pre-check stage should be
\R{checkFunction}s that are applicable to all variable classes. The
default pre-checks, the functions \R{isKey()} and \R{isEmpty()}, check
whether a variable has unique values for all observations or only a
single value for all observations, respectively. If one of these
statements are true, the variable will not be subjected to the SVC
steps.  We can allow empty variables to move on to the SVC step by
only checking for keys in the pre-check step:

@@

As we saw with the \R{check()} function, the summary can be modified
by providing the relevant \R{XXXSummaries} arguments. Setting the
\R{numericSummaries} argument, we can control the summary output by
providing a vector of function names to run for a particular summary. To
only get the median, minimum and the maximum we set
\R{numericSummaries=c("centralValue", "minMax")}:

@@

This section consists of two parts. First, we describe how customized \R{summaryFunction}s, 
\R{visualFunction}s and \R{checkFunctions} can be made, including an introduction to a 
few convenient tools available in \pkg{cleanR} for aiding SVC-function construction. 
Table \ref{table.functionTypes} serves as an overview of the internals of these three
central function types. Secondly, we turn to a 
worked example of how to use custom made functions in practice. Here,
four new SVC functions are defined and used, both interactively and in \R{clean()}.

\begin{table}[htbp]
\footnotesize
\bgroup
\def\arraystretch{1.8}% 
\begin{tabular}{L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}L{0.2\linewidth}}
& \R{summaryFunction} & \R{visualFunction} & \R{checkFunction} \\
\hline
\vspace{0pt} Input (required) & \R{v} - a variable vector \newline \R{...}  &  \R{v} - a variable vector \newline \R{vnam} - the variable name (as character string) \newline \R{doEval} - a logical (\R{TRUE}/\R{FALSE}) controlling the output type of the function & \R{v} - a variable vector \newline \R{nMax} - an integer (or \R{Inf}), controlling how many problematic values are printed, if relevant \newline \R{...}  \\
Input (optional) &  \R{maxDecimals} - number of decimals printed in outputted numerical values.  & - &  \R{maxDecimals}  - number of decimals printed in outputted numerical values.  \\
Purpose & Describe some aspect of the variable, e.g. a central value, its dispersion or level of missingness. & Produce a distribution plot. & Check a variable for a specific issue and, if relevant, identify the values in the variable that cause the issue. \\
Output (required) & A list with entries: \newline \R{\$feature} - a label for the summary value (as character string); \newline \R{\$result} - the result of the summary (as character string) & A character string with \R{R} code for producing a plot. This code should be standalone, i.e. should include the data if necessary. & A list with entries: \newline \R{\$problem} - a logical identifying whether an issue was found; \newline \R{\$message} - a character string (possibly empty) decribing the issue that was found, properly escaped and ready for use in \R{rmarkdown} \\
Output (recommended) & A \R{summaryResult} object, i.e. an attributed list with entries \R{\$feature}, \R{\$result} and \R{\$value}, the latter being the values from \R{\$result} in their original format). & \bgroup
\def\arraystretch{1.8}% 
\hspace*{-.2cm}\begin{tabular}[t]{p{0.45\linewidth} | p{0.45\linewidth}} \vspace*{-1cm}If \R{doEval} is \R{TRUE}: \newline A plot that is opened by the graphic device in \R{R}. & \vspace*{-1cm}If \R{doEval} is \R{FALSE}: \newline   A text string with \R{R} code, as described above. \end{tabular} \egroup  & A \R{checkResult} object, i.e. an attributed list with entries \R{\$problem}, \R{\$message} and \R{\$problemValues}, the latter being either \R{NULL} or the problem causing values, as they were found in \R{v}, whichever is relevant. \\
Tools available for producing the function & \R{summaryResult()} & --- & \R{messageGenerator()} \newline \R{checkResult()} 
\end{tabular}
\egroup
\caption{Reference information for creating new functions to be used
  as part of the \mbox{summarize-,} visualize-, and check steps. The three
  columns correspond to each of the three function types. %\hl{I think the formatting here is still sort of chaotic. Maybe include horizontal lines everywhere? Is it allowed in JSS? It's impossible to read what's in which row right now.}
}
\label{table.functionTypes}
\end{table}

@@

Adhering to the template above is sufficient for using the freshly
made \R{mySummaryFunction()} in \R{clean()}, but we recommend
furthermore adding the new function to the overview of all summary
functions by converting it to a proper \code{summaryFunction}
object. This is done by calling the \R{summaryFunction()} creator with
the user-defined function as the first argument, and additional arguments
\R{description} (an explanatory text which will be added to the attributes of
the function), and \R{classes} (a vector of variable classes the
user-defined function is intended to be applied to, also stored as an
attribute). In other words, a call on the following form should be made:
\begin{Schunk}
\begin{Sinput}
mySummaryFunction <- summaryFunction(mySummaryFunction,
  description = "[Text describing what the summaryFunction does]",
  classes = c([vector of data types that the function is intended for]))
\end{Sinput}
\end{Schunk}
which adds the new function to the output of an
\code{allSummaryFunctions()} call. 
% One comment should be devoted to
% the two attributes of a \R{summaryFunction}. 
If the \R{description} argument is left unspecified, the name of the
function (which in this case is \R{"mySummaryFunction"}) will be filled in and
stored under the \R{description} attribute. What happens if the \R{classes} argument is
not specified depends on the type of \R{mySummaryFunction}. If
\R{mySummaryFunction} is constructed as an \R{S3} generic function with associated
methods, the call to \R{summaryFunction()} will automatically produce
a vector of the names of the classes for which the function can be
called. If \R{mySummaryFunction} is not an \R{S3} generic and
\R{classes} is left unspecified, the attribute will simply be
left empty. Note that the helper function \R{allClasses()} might be useful
for filling out the \R{classes} argument, as it simply lists all
available classes in \pkg{cleanR}:
\begin{Schunk}
\begin{Sinput}
> allClasses()
\end{Sinput}
\begin{Soutput}
[1] "character" "Date"      "factor"   "integer"   "labelled"  
[6] "logical"  "numeric"  
\end{Soutput}
\end{Schunk}
% \hl{Write something here, don't end paragraph with code. Also, maybe move the allClasses() stuff somewhere else, it doesn't really belong under this header. Not sure where to, though.}

@@

myProbValCheckFunction <- checkFunction(myProbValCheckFunction,
  description = "[Some text describing the checkFunction]",
  classes = c([the data types that this function is intended to be used for])
)
\end{Sinput}
\end{Schunk}
One comment should be devoted to the helper function,
\R{messageGenerator()}.  This function's sole purpose is aiding
consistent styling of all \code{checkFunction} messages. The function
simply pastes together the \code{problemMessage} and the
\code{problemValues}, with the latter being quoted and sorted
alphabetically. If the \R{nMax} argument to \R{messageGenerator()} is
not \R{Inf}, only the first \R{nMax} problem values will be pasted
onto the message, accompanied by a comment about how many problem
values were left out (if any).  Note that printing quotes in
\pkg{rmarkdown} requires an extensive amount of character escaping, so
opting for \code{messageGenerator()} really is the easiest solution.

@@

@@

\section{Rubbing down data cleaning challenges}
\label{sec:specificExamples}
%\hl{Lidt poppet titel, men den gamle ("Using cleanR in specific situations")
%fik mig til flabel at tænke "i modsætning til uspecifikke situationer?". Jeg 
%er dog meget åben for en bedre titel.}

@@

On the other hand, if a pdf document is to be produced, a few extra
lines need to be added to the preamble of the master \pkg{rmarkdown}
document --- otherwise, the two-column layout code will produce an
error. The following is an example of how such a master document
preamble YAML might look like and how the \R{cleanR\_toyData.Rmd} file can
then be included:
{\small
\begin{Verbatim}
---
output: pdf_document
documentclass: report
header-includes:
  - \renewcommand{\chaptername}{Part}
  - \newcommand{\fullline}{\noindent\makebox[\linewidth]{\rule{\textwidth}{0.4pt}}}
  - \newcommand{\bminione}{\begin{minipage}{0.75 \textwidth}}
  - \newcommand{\bminitwo}{\begin{minipage}{0.25 \textwidth}}
  - \newcommand{\emini}{\end{minipage}}
---

```{r, child = 'cleanR_toyData.Rmd'}
```
\end{Verbatim}
}
%\hl{Use proper formatting here. How do we do non-R code?}
